---
title: "**3. Distributions of Data and Quality Control Diagnostics**"
author: "Belinda Cornes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE)

library("kableExtra")
library("knitr")
options(stringsAsFactors = FALSE)
library(data.table) 
library(tidyr)     
library(mclust)     
library(rhdf5)      
library(optparse)
library(dplyr)
library(cluster)
library(readxl)
library(psych)
library(tibble)
library(ggplot2)
library(reshape2)
library(qtl2)
library(qtl)
library(dplyr)

gmap <- read_csv("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/gmap.csv")
pmap <- read_csv("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/pmap.csv")
phenos <- read_csv("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/pheno.csv")
genos <- read_csv("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/geno.csv")
covars <- read_csv("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/pheno_covar.csv")

cross <- qtl2::read_cross2("/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/gm.json")
select_phenos <- colnames(cross$pheno)
df_phenos <- cross$pheno

phenos$id <- rownames(phenos)
covars$id <- rownames(covars)
phenos.covars <- merge(phenos[c("id","R_AVG","L_AVG","Both_AVG","deg_avg_r", "deg_avg_l" ,"deg_avg_b")],covars[c("id","Sex")], by=c("id"), all=T)
##reformat
phenos.covars.lg <- phenos.covars %>% gather(variable, value, -c("id","Sex"))
phenos.covars.1 <- subset(phenos.covars.lg, variable %in% c("R_AVG","L_AVG","Both_AVG"))
phenos.covars.2 <- subset(phenos.covars.lg, variable %in% c("deg_avg_r", "deg_avg_l" ,"deg_avg_b"))

````

### **3.1 Phenotype distribution**
In total we have 225 samples, among them 120 females, 105 males. We use 6 phenotypes for QTL analysis. They are:

+ average right eye axial length (`R_AVG`), 
+ average left eye axial length (`L_AVG`), 
+ average axial length for both eyes (`Both_AVG`), 
+ average right eye axial length for degeneration data (`deg_avg_r`), 
+ average left eye axial length for degeneration data (`deg_avg_l`), 
+ average axial length for both eyes for degeneration data (`deg_avg_b`).

Phenotypic distributions are plotted below. 

```{r Phenotype distribution, fig.height = 6, fig.width = 9.5, fig.align = "center"}

#phenotypes:
#print(colnames(df_phenos))

#Axial Length:
#dev.off()
pairs.panels(df_phenos[,c("R_AVG","L_AVG","Both_AVG")], 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             lm = TRUE
)
#dev.off()

box1 <-  ggplot(data=phenos.covars.1, aes(x=variable, y=as.numeric(value), color=Sex, fill=Sex)) +
         geom_boxplot(position = position_dodge(width=0.9)) +
         #ggtitle(paste0("Values of ",v," [random dataframe: ",r,"]")) +
         #labs(y = v) +
         theme(strip.text.x = element_text(size=13),
               axis.text.x = element_text(size = 13, angle = 0),
               axis.text.y = element_text(size = 13, angle = 0),  
               axis.title.x=element_blank(),
               axis.title.y=element_blank(),
               plot.title = element_text(size = 13, face = "bold",hjust = 0.5),
                  #legend.position = "none"
             )
box1



#Degeneration:
#dev.off()
pairs.panels(df_phenos[,c("deg_avg_r", "deg_avg_l" ,"deg_avg_b")], 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             lm = TRUE
)
#dev.off()

box2 <-  ggplot(data=phenos.covars.2, aes(x=variable, y=as.numeric(value), color=Sex, fill=Sex)) +
         geom_boxplot(position = position_dodge(width=0.9)) +
         #ggtitle(paste0("Values of ",v," [random dataframe: ",r,"]")) +
         #labs(y = v) +
         theme(strip.text.x = element_text(size=13),
               axis.text.x = element_text(size = 13, angle = 0),
               axis.text.y = element_text(size = 13, angle = 0),  
               axis.title.x=element_blank(),
               axis.title.y=element_blank(),
               plot.title = element_text(size = 13, face = "bold",hjust = 0.5),
                  #legend.position = "none"
             )
box2


```
QTL analysis requires variables follow normal distribution, from the above distributions, there are no "bad" candidates for QTL analysis.  


However, we removed any samples that were 3 standard deviations from the mean for a particular phenotype.  These were:

```{r outliers, fig.height = 6, fig.width = 9.5, fig.align = "center"}

#outliers
df_phenos.df <- as.data.frame(df_phenos)
#des.3 <- Hmisc::describe(df_phenos[,c("R_AVG","L_AVG","Both_AVG")]) 
des.1 <- pastecs::stat.desc(df_phenos.df[,c("R_AVG","L_AVG","Both_AVG","deg_avg_r", "deg_avg_l" ,"deg_avg_b")]) 
des.2 <- psych::describe(df_phenos.df[,c("R_AVG","L_AVG","Both_AVG","deg_avg_r", "deg_avg_l" ,"deg_avg_b")]) 

#scale(df_phenos[,c("R_AVG")])

df_phenos.df$out.R_AVG <- ifelse(df_phenos.df[,c("R_AVG")] > (des.1[9,1] + 3*des.1[13,1])  | df_phenos.df[,c("R_AVG")] < (des.1[9,1] - 3*des.1[13,1]), 'Outlier','Keep')
df_phenos.df$out.L_AVG <- ifelse(df_phenos.df[,c("L_AVG")] > (des.1[9,2] + 3*des.1[13,2])  | df_phenos.df[,c("L_AVG")] < (des.1[9,2] - 3*des.1[13,2]), 'Outlier','Keep')
df_phenos.df$out.Both_AVG <- ifelse(df_phenos.df[,c("Both_AVG")] > (des.1[9,3] + 3*des.1[13,3])  | df_phenos.df[,c("Both_AVG")] < (des.1[9,3] - 3*des.1[13,3]), 'Outlier','Keep')
df_phenos.df$out.deg_avg_r <- ifelse(df_phenos.df[,c("R_AVG")] > (des.1[9,4] + 3*des.1[13,4])  | df_phenos.df[,c("R_AVG")] < (des.1[9,4] - 3*des.1[13,4]), 'Outlier','Keep')
df_phenos.df$out.deg_avg_l <- ifelse(df_phenos.df[,c("L_AVG")] > (des.1[9,5] + 3*des.1[13,5])  | df_phenos.df[,c("L_AVG")] < (des.1[9,5] - 3*des.1[13,5]), 'Outlier','Keep')
df_phenos.df$out.deg_avg_b <- ifelse(df_phenos.df[,c("Both_AVG")] > (des.1[9,6] + 3*des.1[13,6])  | df_phenos.df[,c("Both_AVG")] < (des.1[9,6] - 3*des.1[13,6]), 'Outlier','Keep')

bad <- NULL
bad$Mouse.ID <- rownames(df_phenos.df)
bad$R_AVG <- ifelse(df_phenos.df$out.R_AVG =="Outlier", 'XX', '')
bad$L_AVG <- ifelse(df_phenos.df$out.L_AVG =="Outlier", 'XX', '')
bad$Both_AVG <- ifelse(df_phenos.df$out.Both_AVG =="Outlier", 'XX', '')
bad$deg_avg_r <- ifelse(df_phenos.df$out.deg_avg_r =="Outlier", 'XX', '')
bad$deg_avg_l <- ifelse(df_phenos.df$out.deg_avg_l =="Outlier", 'XX', '')
bad$deg_avg_b <- ifelse(df_phenos.df$out.deg_avg_b =="Outlier", 'XX', '')
bad[is.na(bad)] <- ""
bad[bad=='NA'] <- ""
df <- do.call(cbind, bad)
bad <- as.data.frame(df)

badind <- subset(bad, 
         bad$R_AVG == 'XX'|
         bad$L_AVG == 'XX'|  
         bad$Both_AVG == 'XX'|
         bad$deg_avg_r == 'XX'|
         bad$deg_avg_l == 'XX'|
         bad$deg_avg_b == 'XX')


#badind <- bad[bad$no_pheno == 'XX',]

badind[] <- lapply(badind, as.character)
#badind$Thaiss_ID <- ifelse(badind$Thaiss == 994 | badind$Thaiss == 995 | badind$Thaiss == 996 |badind$Thaiss == 997 | badind$Thaiss == 998 | badind$Thaiss == 999, "--", bad$Thaiss_ID)

rownames(badind) <- NULL

badind[] %>% 
   dplyr::mutate(
     R_AVG = ifelse(R_AVG == 'XX',
                  cell_spec(R_AVG, color = 'gray',background = 'gray'),
                  ''),
     L_AVG = ifelse(L_AVG == 'XX',
                  cell_spec(L_AVG, color = 'gray',background = 'gray'),
                  ''),
     Both_AVG = ifelse(Both_AVG == 'XX',
                  cell_spec(Both_AVG, color = 'gray',background = 'gray'),
                  ''),
     deg_avg_r = ifelse(deg_avg_r == 'XX',
                  cell_spec(deg_avg_r, color = 'gray',background = 'gray'),
                  ''),
     deg_avg_l = ifelse(deg_avg_l == 'XX',
                  cell_spec(deg_avg_l, color = 'gray',background = 'gray'),
                  ''),
     deg_avg_b = ifelse(deg_avg_b == 'XX',
                  cell_spec(deg_avg_b, color = 'gray',background = 'gray'),
                  '')
     ) %>%
   kable(escape = F,align = c("ccccccccc"),linesep ="\\hline") %>%
   kable_styling("striped", full_width = F) %>%
   column_spec(1:6, width = "3cm") 



##removing outliers
df_phenos.df$Mouse.ID <- rownames(df_phenos.df)
df_phenos.df$R_AVG[df_phenos.df$out.R_AVG == "Outlier"] <- '' 
df_phenos.df$L_AVG[df_phenos.df$out.L_AVG == "Outlier"] <- '' 
df_phenos.df$Both_AVG[df_phenos.df$out.Both_AVG == "Outlier"] <- '' 
df_phenos.df$deg_avg_r[df_phenos.df$out.deg_avg_r == "Outlier"] <- '' 
df_phenos.df$deg_avg_l[df_phenos.df$out.deg_avg_l == "Outlÿer"] <- '' 
df_phenos.df$deg_avg_b[df_phenos.df$out.deg_avg_b == "Outlier"] <- '' 

#df_phenos.df <- df_phenos.df[c(1:15)]
#df_phenos.df$id <- rownames(df_phenos.df)
write.csv(df_phenos.df[c(22,1:15)],"/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts/pheno_cleaned.csv", row.names=F, quote=F)


```

That is, those that have a grey square were removed for that particular phenotype in the QTL mapping. 

### **3.2 Genotyping quality assessment**
The recombination fraction (RF) plot is presented below. Markers on the same chromosome that are closer to each other are assumed to be linked (yellow color); Markers that are far away from each other on the same chromosome or markers on the different chromosomes are assumed to be independent. The patterns in Figure 2 suggests that the quality of genotyping is reasonably good.

```{r Genotyping quality assessment, fig.height = 6, fig.width = 9.5, fig.align = "center"}

cross <- read.cross("csv", "/Users/corneb/Documents/MyJax/CS/Projects/Pinkney/scripts", "geno_t_pheno_covar.csv")
mapthis <- est.rf(cross)
plotRF(mapthis, alternate.chrid=FALSE, what="rf")

```
